---
title: "NicheNet analysis"
output: html_document
---

```{r}
library(nichenetr)
library(Seurat) # please update to Seurat V4
library(tidyverse)
```

```{r}
seuratObj = readRDS("/storage/holab/linxy/iPSC/day7_preprocess/day7iPSC_wUMAP.RDS")
outputFolder = "/storage/holab/linxy/iPSC/day7_NichNet/"
sampleName = "day7"
DimPlot(seuratObj)
DimPlot(seuratObj, group.by = "treatment")
seuratObj@meta.data %>% head()
```

## Read in NicheNet’s ligand-target prior model, ligand-receptor network and weighted integrated networks:

```{r}
ligand_target_matrix = readRDS(url("https://zenodo.org/record/3260758/files/ligand_target_matrix.rds"))
ligand_target_matrix[1:5,1:5] # target genes in rows, ligands in columns

lr_network = readRDS(url("https://zenodo.org/record/3260758/files/lr_network.rds"))
head(lr_network)

weighted_networks = readRDS(url("https://zenodo.org/record/3260758/files/weighted_networks.rds"))
weighted_networks_lr = weighted_networks$lr_sig %>% inner_join(lr_network %>% distinct(from,to), by = c("from","to"))

head(weighted_networks$lr_sig) # interactions and their weights in the ligand-receptor + signaling network
head(weighted_networks$gr) # interactions and their weights in the gene regulatory network
```

## Perform the NicheNet analysis

Here, we use the RNA assay to get the information of gene expression.

```{r}
## receiver
receiver = "Notochord"     ## Primitive streak
expressed_genes_receiver = get_expressed_genes(receiver, seuratObj, pct = 0.10, assay_oi = "RNA")

background_expressed_genes = expressed_genes_receiver %>% .[. %in% rownames(ligand_target_matrix)]
```

```{r}
## sender
sender_celltypes = c("Endoderm", "Mesoderm")  ## cell numbers in NOTOSB - Cardiac mesoderm is too low, remove Mesoderm

list_expressed_genes_sender = sender_celltypes %>% unique() %>% lapply(get_expressed_genes, seuratObj, 0.10, "RNA") # lapply to get the expressed genes of every sender cell type separately here
expressed_genes_sender = list_expressed_genes_sender %>% unlist() %>% unique()
```

## NT vs NOTO

### Define a gene set of interest:

Calculate the differential expression using standard Seurat Wilcoxon test

```{r}
condition_oi = "NOTOSB"
condition_reference = "NOTO" 
meta_column = "treatment"
```

```{r}
seurat_obj_receiver= subset(seuratObj, idents = receiver)
seurat_obj_receiver = SetIdent(seurat_obj_receiver, value = seurat_obj_receiver[[meta_column]])


DE_table_receiver = FindMarkers(object = seurat_obj_receiver, ident.1 = condition_oi, ident.2 = condition_reference, min.pct = 0.10) %>% rownames_to_column("gene")

geneset_oi = DE_table_receiver %>% filter(p_val_adj <= 0.05 & abs(avg_log2FC) >= 0.25) %>% pull(gene)
geneset_oi = geneset_oi %>% .[. %in% rownames(ligand_target_matrix)]
```

### Define a set of potential ligands

```{r}
ligands = lr_network %>% pull(from) %>% unique()
receptors = lr_network %>% pull(to) %>% unique()

expressed_ligands = intersect(ligands,expressed_genes_sender)
expressed_receptors = intersect(receptors,expressed_genes_receiver)

potential_ligands = lr_network %>% filter(from %in% expressed_ligands & to %in% expressed_receptors) %>% pull(from) %>% unique()
```

### Perform NicheNet ligand activity analysis

```{r}
ligand_activities = predict_ligand_activities(geneset = geneset_oi, background_expressed_genes = background_expressed_genes, ligand_target_matrix = ligand_target_matrix, potential_ligands = potential_ligands)

ligand_activities = ligand_activities %>% arrange(-pearson) %>% mutate(rank = rank(desc(pearson)))
ligand_activities
```

```{r}
best_upstream_ligands = ligand_activities %>% top_n(20, pearson) %>% arrange(-pearson) %>% pull(test_ligand) %>% unique()
```

```{r}
DotPlot(seuratObj, features = best_upstream_ligands %>% rev(), cols = "RdYlBu") + RotatedAxis()
ggsave(str_c(outputFolder, sampleName, "_", condition_oi, "vs", condition_reference, "_ligands_activity.png"), width = 12, height = 8)
```

### Infer receptors and top-predicted target genes

#### Active target gene inference

```{r}
active_ligand_target_links_df = best_upstream_ligands %>% lapply(get_weighted_ligand_target_links,geneset = geneset_oi, ligand_target_matrix = ligand_target_matrix, n = 200) %>% bind_rows() %>% drop_na()

active_ligand_target_links = prepare_ligand_target_visualization(ligand_target_df = active_ligand_target_links_df, ligand_target_matrix = ligand_target_matrix, cutoff = 0.33)

order_ligands = intersect(best_upstream_ligands, colnames(active_ligand_target_links)) %>% rev() %>% make.names()
order_targets = active_ligand_target_links_df$target %>% unique() %>% intersect(rownames(active_ligand_target_links)) %>% make.names()
rownames(active_ligand_target_links) = rownames(active_ligand_target_links) %>% make.names() # make.names() for heatmap visualization of genes like H2-T23
colnames(active_ligand_target_links) = colnames(active_ligand_target_links) %>% make.names() # make.names() for heatmap visualization of genes like H2-T23

vis_ligand_target = active_ligand_target_links[order_targets,order_ligands] %>% t()
```

```{r}
p_ligand_target_network = vis_ligand_target %>% make_heatmap_ggplot("Prioritized ligands","Predicted target genes", color = "purple",legend_position = "top", x_axis_position = "top",legend_title = "Regulatory potential")  + theme(axis.text.x = element_text(face = "italic")) + scale_fill_gradient2(low = "whitesmoke",  high = "purple", breaks = c(0,0.0045,0.0090))
p_ligand_target_network
ggsave(str_c(outputFolder, sampleName, "_", condition_oi, "vs", condition_reference, "_target_genes.png"))
```

#### Receptors of top-ranked ligands

```{r}
lr_network_top = lr_network %>% filter(from %in% best_upstream_ligands & to %in% expressed_receptors) %>% distinct(from,to)
best_upstream_receptors = lr_network_top %>% pull(to) %>% unique()

lr_network_top_df_large = weighted_networks_lr %>% filter(from %in% best_upstream_ligands & to %in% best_upstream_receptors)

lr_network_top_df = lr_network_top_df_large %>% spread("from","weight",fill = 0)
lr_network_top_matrix = lr_network_top_df %>% select(-to) %>% as.matrix() %>% magrittr::set_rownames(lr_network_top_df$to)

dist_receptors = dist(lr_network_top_matrix, method = "binary")
hclust_receptors = hclust(dist_receptors, method = "ward.D2")
order_receptors = hclust_receptors$labels[hclust_receptors$order]
    
dist_ligands = dist(lr_network_top_matrix %>% t(), method = "binary")
hclust_ligands = hclust(dist_ligands, method = "ward.D2")
order_ligands_receptor = hclust_ligands$labels[hclust_ligands$order]

order_receptors = order_receptors %>% intersect(rownames(lr_network_top_matrix))
order_ligands_receptor = order_ligands_receptor %>% intersect(colnames(lr_network_top_matrix))

vis_ligand_receptor_network = lr_network_top_matrix[order_receptors, order_ligands_receptor]
rownames(vis_ligand_receptor_network) = order_receptors %>% make.names()
colnames(vis_ligand_receptor_network) = order_ligands_receptor %>% make.names()
```

```{r}
p_ligand_receptor_network = vis_ligand_receptor_network %>% t() %>% make_heatmap_ggplot("Ligands","Receptors", color = "mediumvioletred", x_axis_position = "top",legend_title = "Prior interaction potential")
p_ligand_receptor_network
ggsave(str_c(outputFolder, sampleName, "_", condition_oi, "vs", condition_reference, "_receptors.png"), width = 12, height = 8)
```

### Add log fold change information of ligands from sender cells

```{r}
# DE analysis for each sender cell type
# this uses a new nichenetr function - reinstall nichenetr if necessary!
DE_table_all = Idents(seuratObj) %>% levels() %>% intersect(sender_celltypes) %>% lapply(get_lfc_celltype, seurat_obj = seuratObj, condition_colname = meta_column, condition_oi = condition_oi, condition_reference = condition_reference, expression_pct = 0.10, celltype_col = NULL) %>% reduce(full_join) # use this if cell type labels are the identities of your Seurat object -- if not: indicate the celltype_col properly
DE_table_all[is.na(DE_table_all)] = 0

# Combine ligand activities with DE information
ligand_activities_de = ligand_activities %>% select(test_ligand, pearson) %>% rename(ligand = test_ligand) %>% left_join(DE_table_all %>% rename(ligand = gene))
ligand_activities_de[is.na(ligand_activities_de)] = 0

# make LFC heatmap
lfc_matrix = ligand_activities_de  %>% select(-ligand, -pearson) %>% as.matrix() %>% magrittr::set_rownames(ligand_activities_de$ligand)
rownames(lfc_matrix) = rownames(lfc_matrix) %>% make.names()

order_ligands = order_ligands[order_ligands %in% rownames(lfc_matrix)]
vis_ligand_lfc = lfc_matrix[order_ligands,]

colnames(vis_ligand_lfc) = vis_ligand_lfc %>% colnames() %>% make.names()

p_ligand_lfc = vis_ligand_lfc %>% make_threecolor_heatmap_ggplot("Prioritized ligands","LFC in Sender", low_color = "midnightblue",mid_color = "white", mid = median(vis_ligand_lfc), high_color = "red",legend_position = "top", x_axis_position = "top", legend_title = "LFC") + theme(axis.text.y = element_text(face = "italic"))
p_ligand_lfc
ggsave(str_c(outputFolder, sampleName, "_", condition_oi, "vs", condition_reference, "_lfc_sender.png"))
```

#### Summary visualizations of the NicheNet analysis

```{r}
# ligand activity heatmap
ligand_pearson_matrix = ligand_activities %>% select(pearson) %>% as.matrix() %>% magrittr::set_rownames(ligand_activities$test_ligand)

rownames(ligand_pearson_matrix) = rownames(ligand_pearson_matrix) %>% make.names()
colnames(ligand_pearson_matrix) = colnames(ligand_pearson_matrix) %>% make.names()

vis_ligand_pearson = ligand_pearson_matrix[order_ligands, ] %>% as.matrix(ncol = 1) %>% magrittr::set_colnames("Pearson")
p_ligand_pearson = vis_ligand_pearson %>% make_heatmap_ggplot("Prioritized ligands","Ligand activity", color = "darkorange",legend_position = "top", x_axis_position = "top", legend_title = "Pearson correlation coefficient\ntarget gene prediction ability)") + theme(legend.text = element_text(size = 9))
```

```{r}
# ligand expression Seurat dotplot
order_ligands_adapted = order_ligands
#order_ligands_adapted[order_ligands_adapted == "H2.M3"] = "H2-M3" # cf required use of make.names for heatmap visualization | this is not necessary if these ligands are not in the list of prioritized ligands!
#order_ligands_adapted[order_ligands_adapted == "H2.T23"] = "H2-T23" # cf required use of make.names for heatmap visualization | this is not necessary if these ligands are not in the list of prioritized ligands!
rotated_dotplot = DotPlot(seuratObj %>% subset(seurat_clusters %in% sender_celltypes), features = order_ligands_adapted, cols = "RdYlBu") + coord_flip() + theme(legend.text = element_text(size = 10), legend.title = element_text(size = 12)) # flip of coordinates necessary because we want to show ligands in the rows when combining all plots
```

```{r}
figures_without_legend = cowplot::plot_grid(
  p_ligand_pearson + theme(legend.position = "none", axis.ticks = element_blank()) + theme(axis.title.x = element_text()),
  rotated_dotplot + theme(legend.position = "none", axis.ticks = element_blank(), axis.title.x = element_text(size = 12), axis.text.y = element_text(face = "italic", size = 9), axis.text.x = element_text(size = 9,  angle = 90,hjust = 0)) + ylab("Expression in Sender") + xlab("") + scale_y_discrete(position = "right"),
  p_ligand_lfc + theme(legend.position = "none", axis.ticks = element_blank()) + theme(axis.title.x = element_text()) + ylab(""),
  p_ligand_target_network + theme(legend.position = "none", axis.ticks = element_blank()) + ylab(""),
  align = "hv",
  nrow = 1,
  rel_widths = c(ncol(vis_ligand_pearson)+6, ncol(vis_ligand_lfc) + 7, ncol(vis_ligand_lfc) + 8, ncol(vis_ligand_target)))

legends = cowplot::plot_grid(
    ggpubr::as_ggplot(ggpubr::get_legend(p_ligand_pearson)),
    ggpubr::as_ggplot(ggpubr::get_legend(rotated_dotplot)),
    ggpubr::as_ggplot(ggpubr::get_legend(p_ligand_lfc)),
    ggpubr::as_ggplot(ggpubr::get_legend(p_ligand_target_network)),
    nrow = 1,
    align = "h", rel_widths = c(1.5, 1, 1, 1))

combined_plot = cowplot::plot_grid(figures_without_legend, legends, rel_heights = c(10,5), nrow = 2, align = "hv")
combined_plot
ggsave(str_c(outputFolder, sampleName, "_", condition_oi, "vs", condition_reference, "_all.png"), width = 16, height = 16)
```

### Inferring ligand-to-target signaling paths

```{r}
weighted_networks = readRDS(url("https://zenodo.org/record/3260758/files/weighted_networks.rds"))
ligand_tf_matrix = readRDS(url("https://zenodo.org/record/3260758/files/ligand_tf_matrix.rds"))

lr_network = readRDS(url("https://zenodo.org/record/3260758/files/lr_network.rds"))
sig_network = readRDS(url("https://zenodo.org/record/3260758/files/signaling_network.rds"))
gr_network = readRDS(url("https://zenodo.org/record/3260758/files/gr_network.rds"))
```

```{r}
ligands_all = "FGF2" # this can be a list of multiple ligands if required
targets_all = vis_ligand_target[ligands_all,which(vis_ligand_target[ligands_all,] > 0.004)] %>% names()

active_signaling_network = get_ligand_signaling_path(ligand_tf_matrix = ligand_tf_matrix, ligands_all = ligands_all, targets_all = targets_all, weighted_networks = weighted_networks)

# For better visualization of edge weigths: normalize edge weights to make them comparable between signaling and gene regulatory interactions
active_signaling_network_min_max = active_signaling_network
active_signaling_network_min_max$sig = active_signaling_network_min_max$sig %>% mutate(weight = ((weight-min(weight))/(max(weight)-min(weight))) + 0.75)
active_signaling_network_min_max$gr = active_signaling_network_min_max$gr %>% mutate(weight = ((weight-min(weight))/(max(weight)-min(weight))) + 0.75)

graph_min_max = diagrammer_format_signaling_graph(signaling_graph_list = active_signaling_network_min_max, ligands_all = ligands_all, targets_all = targets_all, sig_color = "indianred", gr_color = "steelblue")

# To render the graph: uncomment following line of code
DiagrammeR::render_graph(graph_min_max, layout = "tree", width = 2400, height = 1600)
DiagrammeR::export_graph(graph_min_max, file_name = str_c(outputFolder, sampleName, "_", condition_oi, "vs", condition_reference, "_graph.png"))
```

```{r}
data_source_network = infer_supporting_datasources(signaling_graph_list = active_signaling_network,lr_network = lr_network, sig_network = sig_network, gr_network = gr_network)
head(data_source_network) 
## # A tibble: 6 x 5
##   from  to    source            database       layer     
##   <chr> <chr> <chr>             <chr>          <chr>     
## 1 SMAD1 TGFBI regnetwork_source regnetwork     regulatory
## 2 SMAD1 TGFBI Remap_5           Remap          regulatory
## 3 SMAD2 LAMC2 harmonizome_CHEA  harmonizome_gr regulatory
## 4 SMAD2 TGFBI harmonizome_CHEA  harmonizome_gr regulatory
## 5 SMAD2 TNC   harmonizome_CHEA  harmonizome_gr regulatory
## 6 SMAD2 TNC   regnetwork_source regnetwork     regulatory
```

