---
title: RT-qPCR Figures 1B and supplementary
author: Antoine Humeau
date: 2023-06-22 09:27:16
output: 
  html_document:
    code_folding: hide
    fig_caption: yes
    highlight: tango
    number_section: yes
    theme: cosmo
    toc: yes
    toc_float: yes
    toc_depth: 4
    keep_md: no
    df_print: paged
---
<style>
.main-container {width: 80%;max-width: inherit;}
</style>

# load

```{r setup}
source(file.path("..", "R_functions", "analyse_prelude.R"))

input_data_folder <- file.path("..", "data","data_valid")
input_data_name <- "ipspine_data_no_SB_20230610_clean.csv"
input_data_path <- file.path(input_data_folder, input_data_name)
```

The data to analyses are `r input_data_name` in the `r input_data_folder` folder. 

```{r load-data}
dt_raw <- data.table::fread(input_data_path, stringsAsFactors = T)

ordre_clone <- c("THERMO F" = "hiPSC a","190 PBMC" = "hiPSC b", "THERMO F NEW" = "hiPSC c",  "CLONE A" = "hiPSC d", "MIPS203.B11" = "hiPSC e", "H1" = "H1")

dt_raw[, clone := factor(clone, levels = names(ordre_clone), labels = ordre_clone)]

logi_in_F <- colnames(dt_raw)[which(dt_raw[,lapply(.SD, is.logical)] == T)]
for (k in logi_in_F) dt_raw[, c(k) := as.factor(get(k))]

gene_vec <- sort(levels(dt_raw$gene))

outcomes_vec <- c("ct", "delta_ct", "fge", "log2_fge")
exposures_vec <- c("gene", "exp", "day", "clone", "NOTO_mRNA")

for (ou in outcomes_vec) {
   tmp_dt <- subset_raw(dt_raw, cible = ou)
   assign(paste0("dt_tidy_", ou), tmp_dt)
}
rm(tmp_dt)
```

The followings variables are outcomes: `r combine_words(outcomes_vec)`.
The followings variables are explanatory variables: `r combine_words(exposures_vec)`.


# Outcome distribution by experimental variables
```{r, results = "asis"}

call_vioplot_1outcome <- function(out) {
   prefix <- out
   res <- knitr::knit_expand(file = file.path("templates","noSB_treatment","ipspine_noSB_treatment_20230610_analyses_template_vioplot_1outcome.Rmd"), out = out, prefix = prefix)
   res2 = knitr::knit_child(text = res, quiet = TRUE)
   cat(res2, sep = '\n')
}
cache = lapply(outcomes_vec, call_vioplot_1outcome)
```


# Fold-change at day x relative to day 0 by gene and clone for NOTO_mRNA

The data used only NOTO_mRNA data.

```{r {{prefix}}-bxp-gene-day-clone-param}
theoutcome <- "log2_fge"

genes_by_outcome <- lapply(theoutcome, function(k){
   tmp_dt <- get(paste0("dt_tidy_", k))
   find_gene <- tmp_dt[,.N, gene][N > 0, sort(as.character(gene))]
   
   find_gene <- tmp_dt[NOTO_mRNA == TRUE & !is.na(get(k)) & day == 7,.N, gene][N > 0, sort(as.character(gene))]
})
names(genes_by_outcome) <- theoutcome

boxplot_by_outcome_gene_cap <- unlist(sapply(theoutcome, function(k){
   intro <- paste0("**", str_replace_all(str_to_upper(k), "_", " "))
   genes <- genes_by_outcome[[k]]
   outro <- paste0("** boxplots depending on clone for NOTO_mRNA transfected cells. Number of data are above the boxes.")
   
   paste0(intro, " and ", genes, outro )
}))
```

```{r {{prefix}}-bxp-gene-day-clone-function}
boxplot_1gene_f <- function(dt, thegene, outcome, main = T, coltitle = "black", ylab = outcome){
   
   thedt <- dt[gene == thegene & !is.na(get(outcome))]
   thedt[, clone := droplevels(clone)]
   
   means <- thedt[, list(N = .N, 
                         mean = mean(get(outcome), na.rm = T),
                         sd = sd(get(outcome), na.rm = T)), by = clone]
   means[N > 1, t95 := qt(p = 0.975, df = N - 1)]
   means[, ic_lower := mean - t95 * (sd/sqrt(N)) ]
   means[, ic_upper := mean + t95 * (sd/sqrt(N)) ]
   means[, x := as.integer(clone)]
   
   yrange <- range(c(thedt[, get(outcome)]), na.rm = T)
 
   if (yrange[1] > log(100, base = 2)) yrange[1] <- log(100, base = 2)
   if (yrange[2] < log(100, base = 2)) yrange[2] <- log(100, base = 2)
   
   
   set.seed(1)
   thebox <- stripchart(get(outcome) ~ clone, data = thedt , main = "", method = "jitter", vertical = T, ylab  = ylab, cex = 1, ylim = yrange, las = 2, pch = 1, xaxt = "n", jitter = 0.1)    
   
   axis(side = 1, at = means$x, labels = paste0(means$clone, " (", means$N, ")"), las = 2)
   
   abline(h = log(100, base = 2), col = "grey", lty = 2)

   points(x = means$clone, y = means$mean, col = "#0570b0", pch = 16, cex = 1 )

   if (main == T) title(main = paste(thegene), outer = F, col.main = coltitle)
   means   
}
```

## RT-qPCR figure sup

```{r {{prefix}}-bxp-gene-day-allclone, fig.cap  = "Fold-change relative to day 0 by gene (panels) and clone (x axis, (Number of data)). blue filled points are the means.", fig.asp =2, results="asis"}
col_pluri <-  "red4"
col_wnt <- "red2"
col_meso <- "green4"
col_endo <- "orange"
col_prim <- "pink"
col_emt <- "purple"

genes_paper <- c( 
                 LEF1 = col_wnt, NODAL = col_wnt, LEFTY1 = col_wnt,
                 NOTO_SYBR = col_meso, TBXT = col_meso, SHH = col_meso,
                 NOGGIN = col_meso, SOX9 = col_meso, 
                 FOXA2 = col_endo, SOX17 = col_endo,
                 TBX6 = col_meso,
                 POU5F1 = col_pluri, NANOG = col_pluri
                  )

autolayout(x = genes_paper, nc = 3, byrow = T)
par(thepar)
par(mar = c(4.5,1.5, 2,.5), cex.axis = .5, oma = c(1,1,0,0), cex.main = .7)

sumfc7 <- lapply(1:length(genes_paper), function(k){
   ge = names(genes_paper)[k]
   thecol = genes_paper[k]
   boxplot_1gene_f(dt = dt_tidy_log2_fge[NOTO_mRNA == T  & day == 7 & !is.na(log2_fge) & gene == ge], thegene = ge, outcome = "log2_fge", main = T, coltitle = thecol, ylab = "")      
})
mtext(side = 2, text = "Fold-Change relative to day 0", outer = T)
mtext(side = 1, text = "Clone", outer = T)
```

## Tables of students test

```{r student-tables, results="asis", eval = T}
print_student <- lapply(1:length(sumfc7), function(k) {
   thetable <- copy(sumfc7[[k]])
   thetable[, t95 := round(t95, 2)]

   wrap_dt(thetable, caption = paste("summary of previous figure", names(genes_paper[k]) ))
          })
htmltools::tagList(print_student) 
```

## RT-qPCR figure 1B

```{r {{prefix}}-bxp-stem, fig.cap = "Mean relative gene expression by gene in day 2 (blue) or day 7 (green). Error bar is one standard error of the mean"}
genes_paper2 <- c(LEF1 = col_wnt, NODAL = col_wnt, LEFTY1 = col_wnt,
                 NOTO_SYBR = col_meso, TBXT = col_meso, SHH = col_meso,
                 NOGGIN = col_meso, SOX9 = col_meso, 
                 FOXA2 = col_endo, SOX17 = col_endo,
                 TBX6 = col_meso
                  )

fig1b_gene <- c("LEF1", "NODAL", "NOTO_SYBR", "TBXT", "SHH", "NOGGIN", "SOX9", "FOXA2", "SOX17", "TBX6" )

dt1b <- dt_tidy_log2_fge[gene %in% fig1b_gene & exp %in% c("20_11", "20_14", "20_16", "21_11", "21_14", "21_19", "22_09", "22_12", "22_20", "22_21") & ( day == 2 | (day == 7 & NOTO_mRNA == T)), 
                         list(Ndata = length(which(!is.na(log2_fge))),
                              Nclone = length(unique(clone)),
                              min = min(log2_fge, na.rm = T),
                              max = max(log2_fge, na.rm = T),
                              value = mean(log2_fge, na.rm = T),
                              sd = sd(log2_fge, na.rm = T)),
                                 .(gene, day)][order(factor(gene, fig1b_gene), day)]

dt1b[, se := sd/sqrt(Ndata)]
dt1b[, xco := as.numeric(factor(gene, fig1b_gene)) + c(-0.2, .2)]

layout(1)
par(thepar)
plot(1, 1, xlim = range(dt1b$xco), ylim = c(0, max(dt1b[, value + se], dt1b[,  value - se])), type = "n", xlab = "gene", ylab = "Relative gene expression / GAPDH (log 2)", xaxt = "n", yaxs = "i")

abline(h = log(100, base = 2), col = "grey", lty = 2)

rect(xleft = dt1b$xco - 0.15, xright = dt1b$xco + 0.15, ybottom = 0, ytop = dt1b$value, col = c("blue2", "green3"))
arrows(x0 = dt1b$xco, y0 = dt1b[, value - se], y1 = dt1b[, value + se], code = 3, angle = 90, length = 0.05, lwd= 1)

cache = sapply(1:length(fig1b_gene), function(k) {
   axis(side = 1, at = k, fig1b_gene[k], col.axis = genes_paper2[fig1b_gene[k]], cex.axis = .5)})
```

```{r {{prefix}}-tbl_stem, results = "asis"}
dt1b2print <- copy(dt1b)
colnames(dt1b2print)[colnames(dt1b2print) == "value"] <- "mean"
wrap_dt(dt1b2print, caption =  "summary by gene and day of qPCR.  sd is standard deviation and se standard error of the mean. xco is the x coordinate in the figure")

```



