---
title: Cleaning of ipspine raw data for triple transfection of October 03 2023
author: Antoine Humeau
date: '`r Sys.time()`'
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_dir = "../reports") }) 
output:
   html_document:
     code_folding: hide
     fig_caption: yes
     highlight: tango
     number_section: yes
     theme: cosmo
     toc: yes
     toc_float: yes
     keep_md: no
     df_print: paged
---

<style>
.main-container {
    width: 80%;
    max-width: inherit;
}
</style>

```{r setup, results = "hide"}
source(file.path("..", "R_functions", "script_raw2transform_prelude.R"))
```

# Explanation

```{r path-parameters}
input_data_folder <- file.path("..", "data","data_raw")
input_data_name <- "ipspine_data_3trans_20231003.xlsx"
input_data_path <- file.path(input_data_folder, input_data_name)

output_data_folder <- file.path("..", "data", "data_valid")
output_data_name <- str_replace(input_data_name, ".xlsx", "_clean.csv")
output_data_path <- file.path(output_data_folder, output_data_name)
```

The document transform the raw data `r input_data_name` in the folder `r input_data_folder` into a clean valid data `r output_data_name` in the folder `r output_data_folder` with `;` as column separator. 

# Load data
```{r load, result = "hide", message = FALSE}
names_sheets <- readxl::excel_sheets(input_data_path)
names_sheets_clean <- str_to_lower(names_sheets) %>%
   str_replace_all(" ", "_")

na_char <- c("", "_", "N/A")
for (x in 1:length(names_sheets)) {
   tmp <- data.table(readxl::read_excel(input_data_path, sheet = names_sheets[x], col_names = F, na = na_char))
   assign(x = paste0("dt_raw_", names_sheets_clean[x]), tmp)
}
rm(tmp)
```

The first sheet (`r knitr::combine_words(names_sheets)`) contain the data and the second sheet metadata
The empty, "_" and "N/A" cells are dealt as NA. The sheet has sub-datasets begin by the columns names (Identifiers of samples and gene names).

# Extract and clean excel file

```{r cleaning-param}
clone_names <- c("MIPS203.B11" = "MIPS"
               )
delta_ct_control_day <- 0 

dico_clean <- rbindlist(list(
   list(variable = "id_replicate", explanation = "Unique identifier of an experiment (replicate and experimental condition)", values = "concatenation of exp, clone, day, NOTO_mRNA, CHIr, sb, sb_day, sb_conc, replicate and extra_replicate"),
   list(variable = "id_experiment", explanation = "Unique identifier of an experimental conditio (same for 2 replicate)", values = "concatenation of exp, clone, day, NOTO_mRNA, CHIr, sb, sb_day, sb_conc"),
   list(variable = "gene", explanation = "The targeted gene measure by qPRC", format = "Capital lellers for gene name"),
   list(variable = "exp_clone", explanation = "concatenation of exp and clone variable. use for extract the delta_ct_control_value"),
    list(variable = "clone", explanation = "Clone use for the experiment. Impute missing clone by the other data if only one clone in the sub-dataset", levels = as.character(combine_words(unique(names(clone_names)), sep = " or ", and = ""))),
   list(variable = "Echantillon", explanation = "Id of experiment without the replicate detail in the raw data"),
   list(variable = "replicate", explanation = "replicate of experimental condition", levels = "1 or 2"),
   list(variable = "exp", explanation = "experimental number in format yy_nn with yy the year of experiment and  nn the number of experiment in the year", format = "yy_nn"),
   list(variable = "exp_year", explanation = "year of the experiment, from exp_year", format = "yyyy"),
   list(variable = "exp_number", explanation = "number of the experiment in the year, from exp_year", format = "positive integer"),
   list(variable = "day", explanation = "Day of qPCR measurerement after beginning of experimnent", levels = "0 or 2 or 3 or 5 or 7"),
   list(variable = "NOTO_mRNA", explanation = "do NOTO rRNA transfection cell between the qPCR of days 2 and 3", levels = "TRUE or FALSE"),
   list(variable = "CHIR", explanation = "do added 3 microMolar of CHIR", levels = "TRUE or FALSE"),
   list(variable = "sb", explanation = "do inject sb molecule", levels = "TRUE or FALSE"),
   list(variable = "sb_day", explanation = "Day of SB injection", levels = "1 or 2 or 3 or 5. NA for non sb condition. Impute missing day by 1."),
   list(variable = "transfection_nb", explanation = "Number of transfection", levels = "1 or 2 or 3. NA for non sb condition"),
   list(variable = "sb_conc", explanation = "concentration of SB injected in microMolar", levels = "5 or 10. NA for non sb condition. Impute missing data by 5."),
   list(variable = "extra_replicate", explanation = "extra text to differenciate several run of the same experiment", levels = "A or B or C or D or E or 0 or 1 or 2 or 3 or 4 or 5 with PF at the begining in some cases"),
   list(variable = "ct", explanation = "Cycle of threshold measure the relative concentration of the gene in the PCR", values = "ct is in range[1,40["),
   list(variable = "ct_control_gene", explanation = "name of control gene of the experiment", levels = "GAPDH_SYBR for SYBR genes and GAPH (glyceraldehyde 3-phosphate dehydrogenase) for other genes"),
   list(variable = "ct_control_replicate", explanation = "ct value for the control gene of the experiment for the replicate"),
   list(variable = "cause_na", explanation = "cause for missing ct value among unmeasured gene in the experiment or out-of-range ct values (≥ 40)", levels = "unmeasure or out-of-range. NA if the data is available. The missing combination of gene x experiment are for unmeasure missing data."),
   list(variable = "delta_ct", explanation = "difference between ct of gene of ct of control gene", values = "ct - ct_control"),
   list(variable = "delta_ct_control", explanation = "control delta_ct for the experiment and gene ", values = "mean of 2 replicate of delta_ct at day = 0 and the same gene"),
   list(variable = "delta_delta_ct", explanation = "difference between the delta ct and the control delta ct condition", values = "delta_ct - delta_ct_control"),
   list(variable = "fge", explanation = "Fold gene expression of the gene", value = "2^-delta_delta_ct"),
   list(variable = "log2_fge", explanation = "base 2 logarithm of 100 * fge")
), use.names = T, fill = T)

add_dico_info <- function(var){
   if (var %in% dico_clean$variable) {
      theexplanation <- dico_clean[variable == var, explanation]
      thevalues <- dico_clean[variable == var, values]
      thelevels <- dico_clean[variable == var, levels]
      theformat <- dico_clean[variable == var, format]

      text2print <- paste0(var, ": ", theexplanation, ".")
      
      if (length(thelevels) == 1) {
         if (thelevels != "" & !is.na(thelevels)) text2print <- paste0(text2print, " It can be ", thelevels, ".")   
      }
      
      if (length(thevalues) == 1) {
         if (thevalues != "" & !is.na(thevalues)) text2print <- paste0(text2print, " It is ", thevalues, ".")   
      }
      
      if (length(theformat) == 1) {
         if (theformat != "" & !is.na(theformat)) text2print <- paste0(text2print, " It take the format ", theformat, ".")   
      }
      return(text2print)
   } else {
      stop(paste(var, "is not in the dictionnary of clean data") )
   }
}
```

The cleaning consists on: 

* search rows for column names: begin by "Echantillon|GAPDH". If Echantillon not found, add as first column. 
* extract the column names
* applied on each sub-dataset contains between two rows of column names
  * remove columns with only missing data
  * rename the columns with the correct gene
      * transform gene in capital
      * if detect 2 occurence of same gene, put SYBR after the second
      * if detect SYBR, SYBER, or endo, put SYBR after the gene name
      * replace SNAI1 and SNAIL by SNAIL1
      * replace NOG by NOGGIN
      * replace PECAM by PECAM1
  * impute the missing data of "Echantillon" column with previous data. Return an error if any Echantillon value is not present 2 or 3 times (number of replicate).
  * Transform Echantillon in capital and replace space by `_`.
  * Extract experimental condition variables from Echantillon
    * `r add_dico_info("replicate")`
    * `r add_dico_info("exp")`
    * `r add_dico_info("exp_year")`
    * `r add_dico_info("exp_number")`
    * `r add_dico_info("day")`
    * `r add_dico_info("NOTO_mRNA")`
    * `r add_dico_info("CHIR")`
    * `r add_dico_info("sb")`
    * `r add_dico_info("transfection_nb")`
    * `r add_dico_info("sb_day")`
    * `r add_dico_info("sb_conc")`
    * Return an error if any missing value in any experimental condition variables
    * `r add_dico_info("extra_replicate")`
    * `r add_dico_info("id_replicate")`
    * `r add_dico_info("id_experiment")`
    * `r add_dico_info("exp_clone")`
  * Pivot in long format, one variable for gene and one for ct. 
  
The process is in the cleaning_one_sheet function and applied on sheet.

```{r cleaning-functions}
cleaning_subdataset <- function(subdt, col.names, sheet_name_input) {
   kept_non_missing_col <- which(apply(subdt, 2, function(j) !all(is.na(j))))

   tidy_subdt <- subdt[, ..kept_non_missing_col]
   
   col.names <- str_to_upper(col.names)
   which_sybr <- str_detect(col.names, "(SYBE?R)|(endo)")
   clean_names <- str_remove(col.names, "(SYBE?R)|(endo)") %>%
      str_squish() %>%
      str_remove("^(_)") %>%
      str_replace("É", "E")
   
   clean_names[which_sybr] <- paste0(clean_names[which_sybr], "_SYBR")
   clean_names[clean_names %in% c("SNAI1", "SNAIL")] <- "SNAIL1"
   clean_names[clean_names %in% c("NOG")] <- "NOGGIN"
   clean_names[clean_names %in% c("PECAM")] <- "PECAM1"
   clean_names[clean_names %in% c("T")] <- "TBXT"
   
   if (!"ECHANTILLON" %in% clean_names) clean_names <- c("ECHANTILLON", clean_names)

   count_name <- table(clean_names)
   double_name <- names(count_name[count_name == 2])
   
   if (length(double_name) > 0) {
      for (do in double_name) {
         clean_names[clean_names == do] <- c(do, paste0(do, "_SYBR"))   
      }
   } 
   if (any(count_name > 2)) stop("some name of gene have more than 2 occcurence in sub-dataset, and it is not manages")
   
   colnames(tidy_subdt) <- clean_names[kept_non_missing_col]

   tidy_subdt[, ECHANTILLON := zoo::na.locf(ECHANTILLON, na.rm = F, fromLast = F)]
  
   check_echantillon <- tidy_subdt[, .N, ECHANTILLON][N != 2 & N != 3]
   if (nrow(check_echantillon) > 0) {
      print(check_echantillon)
      stop("The previous table show the Echantillons without 2 or 3 data rows")
      
   }
   
   gene_cols <- setdiff(colnames(tidy_subdt), "ECHANTILLON")

   tidy_subdt[, ECHANTILLON := str_to_upper(ECHANTILLON)]
   tidy_subdt[, ECHANTILLON := str_replace_all(ECHANTILLON, " ", "_")]
 
   extract_echan <- as.integer(as.factor(tidy_subdt[, ECHANTILLON]))
   diff_echan <- c(1, diff(extract_echan)) != 0
   tidy_subdt[, replicate := abs(diff_echan - 1) + 1]
   
   therep <- 1
   for (ind in 2:length(extract_echan)) {
      if (extract_echan[ind] == extract_echan[ind - 1]) {
         therep[ind] <- last(therep) + 1
      } else {
         therep[ind] <- 1
      }
   }
   
   tidy_subdt[, replicate := therep]
  
   tidy_subdt[, exp := str_extract(ECHANTILLON, "^([0-9]{2}-[0-9]{2})")]
   
   tidy_subdt[, exp := str_replace(exp, "-", "_")]
   tidy_subdt[, exp_year := as.integer(paste0("20", str_sub(exp, 1,2)))]
   tidy_subdt[, exp_number := as.integer(str_sub(exp, 4,5))]
   
   tidy_subdt[, day := str_extract(ECHANTILLON, "([TD] ?[02357])")]
   tidy_subdt[, day := as.integer(str_extract(day, "[02357]"))]
  
   tidy_subdt[, clone := str_extract(ECHANTILLON, paste0("(", paste(clone_names, collapse = ")|("),")"))]
   tidy_subdt[is.na(clone), clone := sheet_name_input]

   #tidy_subdt[, clone := factor(clone, levels = clone_names, labels = names(clone_names))]

   tidy_subdt[, NOTO_mRNA := str_detect(ECHANTILLON, "NOTO") ]
   tidy_subdt[, CHIR := str_detect(ECHANTILLON, "3C") ]
   tidy_subdt[, sb := str_detect(ECHANTILLON, "SB") ]
   tidy_subdt[, transfection_nb := str_extract(ECHANTILLON, "[123]X") ]

   tidy_subdt[, transfection_nb := as.integer(str_extract(transfection_nb, "[123]")) ]

   # check extractions
   nonacols <- c("replicate", "exp", "exp_year", "exp_number", "day", "NOTO_mRNA", "CHIR","sb")
   for (nonacol in nonacols) {
      na_rows <- tidy_subdt[, which(is.na(get(nonacol)))]
      if (length(na_rows) > 0) {
         print(tidy_subdt[na_rows, .SD, .SDcols = c("ECHANTILLON", nonacol)])
         warning(paste("The previous table show data without correct extration of", nonacol, "variable"))
      }
   }
   
   tidy_subdt[sb == TRUE, sb_day := as.integer(str_extract(str_extract(ECHANTILLON, "SB.[JDT][1235]"), "[1235]" )) ]
   tidy_subdt[sb == TRUE & is.na(sb_day), sb_day := 1]
   
   tidy_subdt[sb == TRUE, sb_conc := as.integer(str_extract(str_extract(ECHANTILLON, "[+]_*(5|(10)).{0,5}SB"), "(5)|(10)")) ]
   tidy_subdt[sb == TRUE & is.na(sb_conc), sb_conc := 5]
   
   tidy_subdt[, extra_replicate := stringr::str_remove_all(stringr::str_extract(ECHANTILLON, "((_PF)*_[ABCDE012345])$"), "_")]
   
   tidy_subdt[, id_replicate := paste0(exp, "_", clone, "_D", day, "_NOTO", NOTO_mRNA,"_transfection", transfection_nb, "_SB", sb, "_D", sb_day, "_C", sb_conc, "_", replicate, "_", extra_replicate)]
   tidy_subdt[, id_experiment := paste0(exp, "_", clone, "_D", day, "_NOTO", NOTO_mRNA,"_transfection", transfection_nb, "_SB", sb, "_D", sb_day, "_C", sb_conc)]
   
   tidy_subdt[, exp_clone := paste0(exp, "_", clone)]

   tidy_subdt <- data.table(tidyr::pivot_longer(tidy_subdt, cols  = all_of(gene_cols), names_to = "gene", values_to = "ct", values_drop_na = F))
   tidy_subdt[, ct := as.numeric(ct)]
   colnames(tidy_subdt)[colnames(tidy_subdt) == "ECHANTILLON"] <- "Echantillon"
   return(tidy_subdt)
}

cleaning_one_sheet <- function(dtname) {
   dt <- get(dtname)
   Ntot <- nrow(dt)
   colnames_rows <- which(apply(dt, 1, function(i) any(str_detect(i, pattern = "(.chantillon)|(GAPDH)"))))
   
   ranges_subdt <- data.table(start = colnames_rows + 1, end = c(colnames_rows[-1] - 1, Ntot))
   
   colnames_list <- lapply(colnames_rows, function(i) as.vector(na.omit(unlist(dt[i,]))))
   
   subset_dt_clean <- lapply(1:length(colnames_rows), function(k) {
      print(paste("Subset of dataset number", k))
      thedt_rows <- ranges_subdt[k, ] 
   
      out <- cleaning_subdataset(subdt = dt[thedt_rows[,start]:thedt_rows[, end]] , col.names = colnames_list[[k]], sheet_name_input = dtname )
   })
   
   subset_dt_clean_1 <- rbindlist(subset_dt_clean, use.names = T, fill = T)
}
```

```{r cleaning, results= "hide"}
dt_tidy_l <- lapply(names_sheets_clean[1], function(k){
   print(paste("Sheet", k))
   cleaning_one_sheet(paste0("dt_raw_", k))
})
dt_tidy <- rbindlist(dt_tidy_l, use.names = T, fill = T)
```

# Add variables

The following variable are added:

* `r add_dico_info("ct_control_gene")`
* impute missing gapdh with value of other replicate
* `r add_dico_info("ct_control_replicate")`
* `r add_dico_info("cause_na")`
* `r add_dico_info("delta_ct")`
* `r add_dico_info("delta_ct_control")`
* `r add_dico_info("delta_delta_ct")`
* `r add_dico_info("fge")`
* `r add_dico_info("log2_fge")`

```{r add-variable, eval = T}
dt_tidy[, ct_control_gene := ifelse(str_detect(gene, "SYBR"), "GAPDH_SYBR", "GAPDH")]
na_gapdh_ech <- dt_tidy[is.na(ct) & gene == "GAPDH", Echantillon]
dt_tidy[is.na(ct) & gene == "GAPDH", ct := dt_tidy[!is.na(ct) & gene == "GAPDH" & Echantillon %in% na_gapdh_ech][match(na_gapdh_ech, dt_tidy[!is.na(ct) & gene == "GAPDH" & Echantillon %in% na_gapdh_ech, Echantillon]), ct]
]

extract_ct_control_ri <- dt_tidy[gene == ct_control_gene, list(ct_control_replicate = ct), by = .(id_replicate, ct_control_gene)]

error_control_ct <- extract_ct_control_ri[,.N, .(id_replicate, ct_control_gene)][N>1]
if (nrow(error_control_ct) > 0) {
   print(error_control_ct)
   stop("Previous samples have duplicated rows need rename")
}

dt_clean <- merge(dt_tidy, extract_ct_control_ri, by = c("id_replicate", "ct_control_gene"), all = T)

dt_clean[, cause_na := NA_character_]
dt_clean[is.na(ct), cause_na := "unmeasure"]
dt_clean[ct >= 40, cause_na := "out-of-range"]

dt_clean[, delta_ct := ct - ct_control_replicate]

extract_delta_ct <- dt_clean[day == 0, list(delta_ct = mean(delta_ct, na.rm = T)), by = .(exp_clone, gene)]

dt_clean2 <- merge(dt_clean, extract_delta_ct, by = c("exp_clone", "gene"), all = T, suffixes = c("", "_control"))
dt_clean2[, delta_delta_ct := delta_ct - delta_ct_control]
dt_clean2[, fge := 2^(-delta_delta_ct)]
dt_clean2[, log2_fge := log(100*fge, base = 2)]
```

```{r working, eval = F, include = F}
dt_tidy[id_replicate == "19_14_dt_raw_mips_D7_NOTOTRUE_transfection1_SBFALSE_DNA_CNA_1_NA" & gene == "TBXT", ct]
dt_tidy[id_replicate == "19_14_dt_raw_mips_D7_NOTOTRUE_transfection1_SBFALSE_DNA_CNA_1_NA" & gene == "GAPDH", ct]

dt_tidy[exp == "19_14" & day == 7 & NOTO_mRNA == T & gene == "GAPDH" ]
```


# validate the clean data

```{r manual_check, results = "asis"}
if (nrow(dt_tidy) != nrow(dt_clean)) {
   texte_error <- paste("<span style=\"color:red; font-weight:700\">",
                        "Error: the research of ct_control have issue because dt_clean and dt_tidy  have not the same number of rows.",
                        "</span>")
   cat(texte_error)
}

if (nrow(dt_clean) != nrow(dt_clean2)) {
   texte_error <- paste("<span style=\"color:red; font-weight:700\">",
                        "Error: the research of delta_ct_control have issue because dt_clean2 and dt_clean  have not the same number of rows.",
                        "</span>")
   cat(texte_error)
}

check_redondant_gene <- dt_clean2[,.N, .(id_replicate, gene)][N > 1]
if (nrow(check_redondant_gene) > 0) {
   texte_error <- paste("<span style=\"color:red; font-weight:700\">",
                        "Error: the following table shows the genes with repeated measures for the same id, that is normally not possible.",
                        "</span>")
   cat(texte_error)
   wrap_dt(check_redondant_gene, caption = "Gene and Id with several values in the clean data. It is not possible")
   
}

orphan_dico <- setdiff(dico_clean$variable, colnames(dt_clean2))
if (length(orphan_dico) > 0 ) {
    texte_error <- paste("<span style=\"color:red; font-weight:700\">",
                        "Error: the following variables in dictionnary are not in the clean data:", combine_words(orphan_dico), 
                        "</span>")
   cat(texte_error) 
}

orphan_data <- setdiff(colnames(dt_clean2), dico_clean$variable)
if (length(orphan_data) > 0 ) {
    texte_error <- paste("<span style=\"color:red; font-weight:700\">",
                        "Error: the following variables in clean data are not in the  dictionnary:", combine_words(orphan_data), 
                        "</span>")
   cat(texte_error) 
}
```

```{r validate, fig.cap = "Figure of summary of validation rules", eval = T}
rules <- validate::validator(
   "replicate_levels" = replicate %vin% c(1,2, 3),
   "exp_year_levels" = exp_year %vin% c(2019,2020,2021, 2022,2023),
   "clone_levels" = clone %vin% c("dt_raw_mips"),
   "day_levels" = day %in% c(0,2,3,5,7),
   var_bin := var_group(NOTO_mRNA, CHIR),
   "bin_test" = var_bin %vin% c(FALSE, TRUE),
   "ct_range" = in_range(ct, 1, 40),
   "sb_day_levels" = is.na(sb_day),
   "sb_conc_levels" = is.na(sb_conc),
    "sb_false" = sb == FALSE, 
    "cause_na_levels" = cause_na %vin% c("unmeasure", "out-of-range") | is.na(cause_na))

check_tidy <- validate::confront(dt_clean2, rules)

wrap_dt(validate::summary(check_tidy), caption = "Summary of validation rules for tidy data")

validate::plot(check_tidy, fill = c("#d95f02", "#1b9e77", "#7570b3"))
```

# Export data

The exported data have `r nrow(dt_clean2)` rows and `r ncol(dt_clean2)` columns named: `r combine_words(colnames(dt_clean2))`.
```{r tidy-print}
wrap_dt(dico_clean, caption = "Dictionnary of the clean data")

write.table(dico_clean, str_replace(output_data_path, "_clean", "_clean_dictionary"), row.names = F, sep = ";")

write.table(dt_clean2, file.path(output_data_path), row.names = F, sep = ";")
```





