---
title: Cleaning of ipsinie quantification raw data of 27 march 2023
author: Antoine Humeau
date: '`r Sys.time()`'
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_dir = "../reports") }) 
output:
   html_document:
     code_folding: hide
     fig_caption: yes
     highlight: tango
     number_section: yes
     theme: cosmo
     toc: yes
     toc_float: yes
     keep_md: no
     df_print: paged
---

<style>
.main-container {
    width: 80%;
    max-width: inherit;
}
</style>

```{r setup, results = "hide"}
source(file.path("..", "R_functions", "script_raw2transform_prelude.R"))
```

# Explanation

```{r path-parameters}
input_data_folder <- file.path("..", "data","data_raw")
input_data_name <- "ipspine_data_quantification_20230327.xlsx"
input_data_path <- file.path(input_data_folder, input_data_name)

output_data_folder <- file.path("..", "data", "data_valid")
output_data_name <- "ipspine_data_quantification_20230327_clean.csv"
output_data_path <- file.path(output_data_folder, output_data_name)
```

The document transform the raw data `r knitr::combine_words(input_data_name)` in the folder `r input_data_folder` into a clean valid data `r output_data_name` in the folder `r output_data_folder` with `;` as column separator. 

# Load data
```{r load, result = "hide", message = FALSE}
na_char <- c("", "_", "N/A")
dt_raw <- data.table(readxl::read_excel(input_data_path, col_names = T, na = na_char))
colnames(dt_raw) <- str_to_lower(colnames(dt_raw))
```

The empty, "_" and "N/A" cells are dealt as NA. Empty columns and rows are removed. Each sheet has sub-datasets begin by the numero manip.

```{r cleaning, results= "hide"}
dt_clean <- copy(dt_raw)

dt_clean[, gene_ratio_name := str_remove_all(gene_ratio_name, " ")]
dt_clean[str_detect(gene_ratio_name, "FOXA2.{2}SOX17"),  gene_ratio_name := sapply(str_split(gene_ratio_name, "/"), function(k) paste(k[2], k[1], sep = "/"))]
dt_clean[, gene_ratio_name := str_replace(gene_ratio_name, "/", " / ")]

dt_clean[, gene_combo := str_remove_all(gene_ratio_name, "[+-]")]

dt_clean[clone == "Thermo F new", clone := "THERMO F"]
dt_clean[relative_count == "NA", relative_count := NA]
dt_clean[, relative_count := as.numeric(as.character(relative_count))]

dt_clean[, NT := str_detect(echantillon, "NT")]

dt_clean[, day_NT := paste0(day, ifelse(NT == T, "_NT", ""))]
```

# Export data

The exported data have `r nrow(dt_clean)` rows and `r ncol(dt_clean)` columns named: `r combine_words(colnames(dt_clean))`.
```{r tidy-print}
write.table(dt_clean, file.path(output_data_path), row.names = F, sep = ";")
```

